Model-Based and Model-Free Deep Reinforcement Learning in Accelerator Operation with Application to FERMI FEL

### 2020-12-16 main.tex
- paper changes

	\title{Uncertainty Aware Model-Based Reinforcement Learning in Accelerator Operation with Application to FERMI FEL}

	\author{Simon Hirlaender}
	%\thanks{These two authors contributed equally}
	% \affiliation{University of Malta}
	\affiliation{University of Salzburg, Salzburg, Austria}
	\author{Niky Bruchon}%, Gianfranco Fenu, Felice Andrea Pellegrino, Erica Salvato}
	\affiliation{University of Trieste, Trieste, Italy}
	% \author{Gianluca Valentino}
	% \affiliation{University of Malta, Msida, MT}
	% \author{Marco Lonza, Giulio Gaio}
	% \affiliation{Elettra Sincrotrone Trieste, Trieste, IT}
	% \author{Verena Kain}
	% %\thanks{These two authors contributed equally}
	% \email{verena.kain@cern.ch}

@@

	\begin{abstract}
		In this paper, we compare a model-based reinforcement learning to a model-free reinforcement learning method applied to the FERMI FEL system. Both approaches hold tremendous promise, and the principal goal of this paper is to show how reinforcement learning can be utilised on an operational level in a reasonable training time on real accelerator physics problems. In terms of representational power and sample-efficiency, the model-based approach is better, while the asymptotic performance of the model-free method is slightly superior. The model-based algorithm is implemented in a DYNA-style using an uncertainty aware model, and the model-free algorithm is based on tailored deep Q-learning. In both cases, the techniques were adapted to show increased noise robustness as omnipresent in accelerator control problems. Code is released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper}.

@@

	%\tableofcontents
	\section{Introduction and Motivation}
	In particle accelerator operation, one main goal is to provide stable and reproducible performance. Achieving this demands to consider several control problems simultaneously \cite{Kain2020,Scheinker2018}. Especially if there is no way to model the physics a priori, one might use optimisation techniques as, e.g. derivative-free optimisers (DFOs) \cite{Huang2013,Bruchon2017,Scheinker2020,Hirlaender2019,Welsch2015,Albright2019} or model-based optimisations as Gaussian processes \cite{Hanuka2020,Roussel2020} to restore or maintain the performance.\\
	 Recently, the community has begun to explore a novel approach to solve control problems. Reinforcement learning (RL) \cite{Bruchon2020,Bruchon2019,Kain2020,Pang2020} unveils several advantages over optimisation methods:
	\begin{itemize}
		\item It covers a larger class of problems, as RL optimises a sequence of decisions.
		\item It memorises the problem and does not always begin from zero as a DFO.
		\item Already existing data can be used.
		\item Adaption to non-stationary systems is possible.
		\item The underlying structure of the problem might be deduced.
	\end{itemize}
	One demanding aspect using RL is the number of iterations needed to train a controller - the sample-efficiency, since RL methods are known to be data-hungry \cite{Sutton2018,DulacArnold2019}, and a second critical aspect is the robustness of the training in real world problems itself.\\
	In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at \emph{Elettra Sincrotrone Trieste}.\\
	 Two different algorithm classes, a modification of the normalized advantage function \emph{NAF} (\cref{ss:Normalized advantage function}) and a novel anchored ensembling model-based algorithm, the \emph{AE-DYNA} (\cref{ss:Uncertainty aware DYNA-style reinforcement learning}), were applied successfully to the FERMI FEL problem and reveal different advantages. The critical aspects are addressed, and tailored solutions, generally applicable to problems showing a sample-complexity as in accelerator control problems are presented.
	 \subsection{An Overview of the Main Results}
	 We demonstrate optimization techniques that simultaneously adjust 4 parameters, the tilt and incline of 2 mirrors that control the trajectory and overlap of the seed laser in the FERMI FEL relative to the beam in the undulator. We show that our algorithms require at most 1000 training sets, after which they are able to maximize the FERMI FEL’s output intensity to $>$ that 95\% of the maximum within only $\sim 3$ to $\sim 5$ steps depending on which algorithm is used. Also we show the operational applicability of purely model-based reinforcement learning in a accelerator control problem.\\
	  Furthermore, our algorithms can perform real-time learning based on data and therefore can be re-run and re-trained for time-varying systems.
	The paper is organised as follows:
	\begin{itemize}
		\item Description of the problem set-up at FERMI
		\item Overview of RL
		\item Design decisions of the implementations used in these studies and theoretical concerns
		\item Results of the experiments 
		\item Summary and outlook
	\end{itemize}
	% The methods proposed are the iterative Linear Quadratic Regulator (iLQR) - an extension to non-linear systems of the fundamental control theory Linear Quadratic Regulator (LQR) problem -  and the Normalised Advantage Function (NAF) - a continuous variant of the Q-learning algorithm from reinforcement learning field. An exhaustive presentation of both techniques is provided in the following sections.

@@

	For this reason, the proposed study focuses on the control of the seed laser trajectory in the modulator, looking at FERMI performance as a reference.

@@

		\bs_{t+1} = \bs_t+\ba_t.

@@

@@

	After a defined number of training episodes, a verification of 50 episodes is performed. In the second phase, the RL only applies the learned policy.

@@

	\section{Deep Reinforcement learning}
	In what follows, we denote the set of all states $\bs$ by $\states$ and the set of all actions $\ba$ by $\actions$. Relative to state and action vectors, $(\bs,\ba)$, we define a reward function $\reward : \states \times \actions \rightarrow \real$, a scalar $\discount \in (0,1]$ called discount factor and an initial state distribution $\initstate$. $\transitions(\bs_{t+1}|\bs_t,\ba_t)$ characterises the probability to end in a state $\bs_{t+1}$ if an action $\ba_t$ is taken at state $\bs_t$. The tuple \mbox{$\mdp = (\states,\actions,\transitions,\initstate,\reward,\discount)$} defines a Markov decision process.\\
	Goal of reinforcement learning is to find a $\policy^*: \bs\mapsto \ba$, which is the solution of:
		J(\policy^*)=\max J (\policy)  = 
		\mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],\label{eq:cumulative_reward}
	where $p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)$ is the distribution of all trajectories $\traj := (\bs_0, \ba_0, \bs_1, \ba_1,\dots \bs_\horizon,\ba_\horizon)$ drawn by $\policy$ with a horizon $\horizon$.\\
	In the modern field of RL one distinguishes if the policy $\policy(\ba)\approx \policy_\phi(\bs)$ or the state-action value function $Q(\bs,\ba) \approx Q_\theta(\bs,\ba)$ is approximated using a high capacity function approximator, as e.g. a deep neural network.\\
	 In the first case $\policy$ is optimized directly and one speaks about policy gradient methods \cite{Sutton2018, Williams1992,Baxter2011,pmlr-v28-levine13, Schulman2015,Schulman2017}. To train these methods, a large number of direct interactions with the system is required, because they belong to on-policy algorithms. In the latter one deals with approximate dynamic programming, which we shortly treat in \cref{s:Model free reinforcement learning}. One advantage of the latter approach is, that it can use previously stored data as it is an off-policy algorithm and generally needs less data to be trained. A combination of both is also used, known as actor-critic methods \cite{Szepesvari2010,Lillicrap2015,pmlr-v32-silver14}.
	\subsection{Model free reinforcement learning - MFRL}\label{s:Model free reinforcement learning}
	\tocheck{Why Q learning?}
	We focus on the main principles of MBRL. An overview can be found e.g. in \cite{Sutton2018,Levine2020}. In the following, we discuss the normalised advantage function (\emph{NAF}) algorithm in some detail, as it has good characteristics for accelerator control problems as its high sample efficiency\cite{Kain2020,Hirlaender2020a}.

@@

		V^\policy(\bs_t) := \mathbb E_{\traj \sim p_{\policy}(\traj|\bs_t)}\left[\sum_{t=t'}^{\horizon}\discount^{(t'-t)} r(\bs_t,\ba_t)\right],\label{eq:state-value-function}

@@

	The state-action-value function $Q^\policy(\bs,\ba)$, which tells us how good an action $\ba$ in a state $\bs$ following $\policy$ is, in terms of the expected return:

@@

		Q^\policy(\bs_t,\ba_t) := \mathbb E_{\traj \sim p_{\policy}(\traj|\bs_t,\ba_t)}\left[\sum_{t=t'}^{\horizon}\discount^{(t'-t)} r(\bs_t,\ba_t)\right].

@@

	is expressed as $Q^\policy_\theta(\bs,\ba)$, where $\theta$ denotes the parameters of the function approximator, such as the weights of a neural network. By satisfying the Bellmann-optimality equation $Q^\policy_\theta$ can be trained towards the optimal $Q^*(\bs,\ba)$ by minimizing the Bellman error:

@@

		\policy_\theta(\bs)=\delta(\ba-\arg\!\max_\ba Q_\theta(\bs,\ba)).

@@

	The Bellman operator $\bellman^*$ has a unique fixed point but is non-linear, due to the $\max$ - operator \cite{Sutton2018}:

@@

	The nature of this equation can cause overestimation and other complications, when using a function approximator and several attempts exist to overcome the difficulties \cite{Hasselt2015,Mnih2013,Lillicrap2015,Gu2016,Wang2015}.
	\subsubsection{Normalized advantage function}\label{ss:Normalized advantage function}

@@

	If a specific quadratic form of the $Q$ function is assumed \cite{Gu2016}:

@@

		&\mathcal L(\theta)=\\\notag&\left( (\reward(\bs_t,\ba_t)+\gamma \min_{1,2} V_{\theta^i_\text{targ}}(\bs_{t+1}) - (1+\gamma) Q_\theta(\bs_t,\ba_t)\right)^2

@@

	$\theta_\text{targ}$ are the weights of a target network, which is softly updated. To stabilize the network training a small clipped artificial noise is added to the actions called \emph{smoothing} as described in \cref{appendix:naf2}.\\
	 This implementation has an extremely good sample-efficiency for suitable problems as can be found often in accelerators. We use it as the baseline for the FERMI-FEL control problem, as it yields good results. Additional changes to the original proposal \cite{Gu2016} and a previous implementation used for experiments using a prioritized replay buffer \cite{Hirlaender2020a} are discussed in \cref{appendix:naf2}.

@@

	A nice overview of model-based reinforcement learning (MBRL) methods is provided in \cite{Wang2019}. Suitable methods are \cite{Gal2016,6654139}, based on Gaussian processes and Bayesian Neural Networks and back-propagation in the dynamics model or \cite{Chua2018,Wang2019a}, using highly efficient sample based methods (e.g. the cross entropy method \cite{Boer2005}). We focus on model-based data generation.\\
		In MFRL the dynamics $\transitions$ of an environment is learned implicitly, while in model-based reinforcement learning MBRL $\transitions$ is learned explicitly and approximated as $\hat \transitions$. Additionally in our case the reward is included in the model \footnote{In some applications $\reward$ can be deduced from the current state or the state change. In such a situation the modelling can be simplified to model directly $\hat\transitions$}:

@@

	An overview of the used method is shown in \cref{fig:MBRL_overview}. In the beginning, the data is collected or read from previously collected data. An uncertainty aware model of the dynamics is trained, using 'anchored ensembling' \cite{Pearce2018} on the data, which allows taking the aleatoric (measurement errors) as well as the epistemic uncertainty (lack of data) into account. It was motivated by the fact that pure epistemic ensemble techniques as \emph{ME-TRPO} were reported to be very sensitive to aleatoric noise \cite{Wang2019}. Alternatives as \cite{Chua2018,Janner2019,Wang2019a} also take these uncertainties into account using probabilistic ensembles. Our implementation is simpler while showing similar performance.\\
	 Subsequently, an MFRL algorithm is deployed to learn the controller on the learned model by only using the synthetic data by taking the model uncertainty into account, as explained in the next section.\\
	  After a defined number of training steps, called epoch, the controller is validated on each model of the ensemble individually. If there is no improvement in a number smaller than the total number of the models, the controller might be tested on the real environment for several episodes. In this way 'over-fitting' on a wrong model is avoided, as discussed in \cite{Kurutach2018}. The training is stopped if the controller solves the problem on the real environment. Otherwise, a batch of new data is collected and the model is improved and consecutively a new epoch starts.

@@

	\subsection{Critical design decisions}\label{ss:critical_design}
In the following, the most important aspects of a successful application of model-based RL are discussed. We made some algorithmic design decisions, which we think are beneficial for problems typically found in accelerator operation.\\

@@

The 'anchored ensembling' method, as treated in \cref{appendix:The impact of noise}, usually yields good results already with a small number of models. Empirical results showed that already three models were satisfactory to see definite improvements over a single network approach (see \cref{fig:Compare_models_sizes}). The main goal is not to determine the exact posterior probability of the model, but, as already mentioned, providing the uncertainty to not over-train the controller at areas without sufficient data.\\

@@

The prior is controlled by the initialisation of the weights of the network. It defines how large the variation of the approximated function is expected. The weights of the last layer were normalised by the number of its nodes. For the moment, only homoscedastic Gaussian errors are considered with a standard deviation is $\sigma_\epsilon$ is also respected in a regularisation term \cref{appendix:The impact of noise} and improves the training in the presence of noise as can be seen in \cref{fig:comparsion_noise_ae_dyna}. \\
Several methods of network training were implemented. \emph{Early stopping} \cite{Goodfellow2016}, with a learning rate of $10^{-3}$ to $10^{-4}$  was employed  with a waiting time of about 20 steps and a validation ratio of 20\%. The number of training steps is increasing with more data in close-by areas. It is followed by a shrinking of the uncertainty at regions where no data is available, leading to better training performance. A fixed maximal loss threshold was also beneficially tested. In the experiments, a combination of both was taken.\\
One of the most crucial points is how model uncertainties are taken into account by the MFRL controller. Several different approaches were tried. One way is to take the mean of the models $\mu_m$ and the standard deviation $\sigma_m$ and sample from $\gauss(\mu_m,\sigma_m)$. Another strategy is to randomly select a model to provide at each training step, which demonstrated the best performance the original implementation of the \emph{ME-TRPO} algorithm and was used in the experiments labelled \emph{ME-TRPO} \cite{Kurutach2018}. A pessimistic selection would only select the model resulting in the lowest predicted reward and showed constant but too slow improvement for our online-training.
Good results were also obtained by following a randomly selected single model each full episode in combination with the soft-actor-critic (\emph{SAC}) \cite{fujimoto2018addressing,Hill2018}, and were used in the experiments labelled \emph{AE-DYNA-SAC}. \\
The number of data-points in a new batch to improve the model has to be chosen carefully. It is influenced by the number of randomly collected initial data-points (and $\policy$). The initialisation phase has to be selected not to small to minimise chances getting trapped in a local minimum for too long. Afterwards, from our experience, at least one full episode per epoch should be taken. We decided to use a short horizon to diminish the impact of the compound error \cite{Janner2019}, the accumulation error following a wrong model, as well known in DYNA-style approaches. The maximal number of steps in our experimental runs was ten. During the data collected on the real system, the latest learned policy was taken with some small additional Gaussian noise to improve exploration.\\
To decide which MFRL algorithm to use, two main algorithm classes have to be considered: the on and the off-policy algorithms. The online algorithm shows a more stable convergence in general, while off-policy algorithms have the advantage that the data buffer can be filled with real data. An attractive off-policy algorithm is \emph{SAC}. The \emph{SAC} not only tries to maximise $J$ \cref{eq:cumulative_reward} but also simultaneously is regularised to maximise the entropy in the action space \cite{Haarnoja2018a}. In this way, exploration is encouraged to avoid getting stuck in a local optimum and has a good trade-off between exploration and exploitation.\\
On the other hand, the on-policy algorithm TRPO \cite{Schulman2015} provides some theoretical improvement guarantees. As reported in \cite{Kurutach2018} using proximal policy optimisation \cite{Schulman2017} as an online MFRL algorithm does not show the same performance, which was confirmed in our tests. In the latter case, the policy was improved over the whole training. For the \emph{AE-DYNA-SAC} the policy was reset after retraining the model, to profit from its exploration features.

>>>>

	\title{Model-Based and Model-Free Deep Reinforcement Learning in Accelerator Operation with Application to FERMI FEL}

	\author{Simon Hirlaender}\email{simon.hirlaender@sbg.ac.at}
	%\thanks{These two authors contributed equally}
	% \affiliation{University of Malta}
	\affiliation{IDA Lab, University of Salzburg, 5020 Salzburg, Austria}
	\author{Niky Bruchon}%, Gianfranco Fenu, Felice Andrea Pellegrino, Erica Salvato}
	\affiliation{University of Trieste, Trieste, Italy}
	% \author{Gianluca Valentino}
	% \affiliation{University of Malta, Msida, MT}
	% \author{Marco Lonza, Giulio Gaio}
	% \affiliation{Elettra Sincrotrone Trieste, Trieste, IT}
	% \author{Verena Kain}
	% %\thanks{These two authors contributed equally}

@@

	\begin{abstract}
		Reinforcement learning holds tremendous promise in accelerator controls, and the principal goal of this paper is to show how reinforcement learning can be utilised on an operational level on accelerator physics problems.
		We compare a model-based reinforcement to a model-free reinforcement learning method applied to the FERMI FEL system.  In terms of representational power and sample-efficiency, the model-based approach is overtaking, while the asymptotic performance of the model-free method is slightly superior. The model-based algorithm is implemented in a DYNA-style using an uncertainty aware model, and the model-free algorithm is based on tailored deep Q-learning. In both cases, the techniques were adapted to show increased noise robustness as omnipresent in accelerator control problems. Code is released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper}.

@@

	%\tableofcontents
	\section{Introduction and Motivation}
	In particle accelerator operation, one main goal is to provide stable and reproducible performance. Achieving this demands to consider several control problems simultaneously \cite{Kain2020,Scheinker2018}. Especially if there is no way to model the physics a priori, one might use optimisation techniques as, e.g. derivative-free optimisers (DFOs) \cite{Huang2013,Bruchon2017,Scheinker2020,Hirlaender2019,Welsch2015,Albright2019} or model-based optimisations as Gaussian processes \cite{Hanuka2020,Roussel2020} to restore or maintain the performance.\\
	 Recently, the community has begun to explore a novel approach to solve control problems. Reinforcement learning (RL) \cite{Bruchon2020,Bruchon2019,Kain2020,Pang2020} unveils several advantages over optimisation methods:
	\begin{itemize}
		\item It covers a larger class of problems, as RL optimises a sequence of decisions.
		\item It memorises the problem and does not always begin from zero as a DFO.
		\item Already existing data can be used.
		\item Adaption to non-stationary systems is possible.
		\item The underlying structure of the problem might be deduced.
	\end{itemize}
	One demanding aspect using RL in real world applications is the number of iterations needed to train a controller - the sample-efficiency, since RL methods are known to be data-hungry \cite{Sutton2018,DulacArnold2019}. A second critical aspect to be considered is the robustness of the training in a noisy data limited environment.\\
	In this paper, we present the study carried out to solve the maximisation-problem of the radiation intensity generated by a seeded Free-Electron Laser (FEL) on the Free Electron laser Radiation for Multidisciplinary Investigations (FERMI) at \emph{Elettra Sincrotrone Trieste}.\\
	 We successfully applied two different algorithm classes, a highly sample-efficient variant of continuous Q-learning (\cref{ss:Normalized advantage function}) and a newly proposed 
	  model-based algorithm, which we call \emph{AE-DYNA} (\cref{ss:Uncertainty aware DYNA-style reinforcement learning}), to the FERMI FEL problem.\\ The methods reveal different characteristics addressing the stated critical aspects. They are generally applicable to problems showing a sample-complexity as in common accelerator control problems.
	 \subsection{An Overview of the Main Results}
	 We demonstrate optimization techniques that simultaneously adjust 4 parameters, the tilt and incline of 2 mirrors that control the trajectory and overlap of the seed laser in the FERMI FEL relative to the beam in the undulator. We show that our algorithms require significantly less than 1000 training sets, after which they are able to maximize the FERMI FEL’s output intensity to $>$ that 95\% of the maximum within only $\sim 3$ to $\sim 5$ steps depending on which algorithm is used. Also we show the operational applicability of purely model-based reinforcement learning in an accelerator control problem.\\
	  Furthermore, our algorithms can perform real-time learning based on data and therefore can be re-run and re-trained for time-varying systems.\\
	The paper is organised as follows:
	\begin{itemize}
		\item Description of the problem set-up at FERMI
		\item Overview of RL
		\item Design decisions of the implementations used in these studies and theoretical concerns
		\item Results of the experiments 
		\item Discussion, outlook and summary
	\end{itemize}
	% The methods proposed are the iterative Linear Quadratic Regulator (iLQR) - an extension to non-linear systems of the fundamental control theory Linear Quadratic Regulator (LQR) problem -  and the Normalised Advantage Function (NAF) - a continuous variant of the Q-learning algorithm from reinforcement learning field. An exhaustive presentation of both techniques is provided in the following sections.

@@

%	For this reason, the proposed study focuses on the control of the seed laser trajectory in the modulator, looking at FERMI performance as a reference.

@@

		\bs_{t+1} = \bs_t+\ba_t,

@@

	and is limited by an $ |\ba_t|\leq \ba_\text{max}=\frac{1}{12}$.

@@

	After a defined number of training episodes, a verification of 50 or 100 episodes is performed. In the second phase, the RL only applies the learned policy.

@@

	\section{Deep Reinforcement learning}
    Most commonly two main categories of deep RL algorithms are distinguished: model-free RL (MFRL) and model-based RL (MBRL). MFRL algorithms learn directly from the interaction with a system, while MBRL algorithm learn a dynamics model from these interactions first. Despite the success of MFRL in several domains as robotics and video games \cite{Heess2017, Schulman2017,Silver2014,Lillicrap2015,OpenAI2018}, there applications is mainly restricted to simulated environments due to their low sample-efficiency. Yet putting constrains on the algorithms, as discussed in \cref{s:Model free reinforcement learning}, partly cures the sample inefficiency turning them into applicable tools for real world problems.\\
      On the other hand MBRL showing a much better sample efficiency. The challenge for these methods is to learn an accurate model, which turns out to be rather hard and leads to the so called model-bias problem \cite{Deisenroth2011}. In \cref{ss:Uncertainty aware DYNA-style reinforcement learning} we discuss attempts to capture the model uncertainty to alleviate this obstacle by making use of model ensembles and Bayesian neural networks resulting in algorithms, exhibiting a high sample-efficiency and a low model-bias.\\
	In what follows, we denote the set of all states $\bs$ by $\states$ and the set of all actions $\ba$ by $\actions$. Relative to state and action vectors, $(\bs,\ba)$, we define a reward function $\reward : \states \times \actions \rightarrow \real$, a scalar $\discount \in (0,1]$ called discount factor and an initial state distribution $\initstate$. $\transitions(\bs_{t+1}|\bs_t,\ba_t)$ characterises the probability to end in a state $\bs_{t+1}$ if an action $\ba_t$ is taken at state $\bs_t$. The tuple \mbox{$\mdp = (\states,\actions,\transitions,\initstate,\reward,\discount)$} defines a Markov decision process.\\
	Goal of reinforcement learning is to find a $\policy^*: \bs\mapsto \ba$, which is the solution of:
	\begin{align}
		J(\policy^*)=\max J (\policy)  = 
		\mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],\label{eq:cumulative_reward}
	\end{align}
	where
	\begin{align}
	p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)
	\label{eq:trajectory_distribution}
	\end{align}

is the distribution of all trajectories $\traj := (\bs_0, \ba_0, \bs_1, \ba_1,\dots \bs_\horizon,\ba_\horizon)$ drawn by $\policy$ with a horizon $\horizon$.

	\subsection{Model free reinforcement learning}\label{s:Model free reinforcement learning}
		In the modern field of RL one differentiates if the policy $\policy(\ba)\approx \policy_\phi(\bs)$ or the state-action value function $Q(\bs,\ba) \approx Q_\theta(\bs,\ba)$ is approximated using a high capacity function approximator, as e.g. a deep neural network.\\
	In the first case $\policy$ is optimized directly and one speaks about policy gradient methods \cite{Sutton2018, Williams1992,Baxter2011,pmlr-v28-levine13, Schulman2015,Schulman2017}. To train these methods, a large number of direct interactions with the system is required, since they belong to \emph{on-policy} algorithms. Hence, we rule them out for direct online training but make use of their characteristics in MBRL (\cref{ss:critical_design}).\\
	 In the second case we speak about Q-learning, which belongs to approximate dynamic programming. One advantage is the possible usage of use previously stored data as we use it in an \emph{off-policy} fashion, which generally needs less data to be trained than \emph{on-policy} methods. A combination of both is also used, known as actor-critic methods \cite{Szepesvari2010,Lillicrap2015,Silver2014}.\\
	 As already adduced, MFRL algorithms learn directly from interactions with the system. An exhaustive overview can be found e.g. in \cite{Sutton2018,Levine2020}. In the following, we discuss the normalised advantage function (\emph{NAF}) algorithm \cite{Gu2016} in some detail, as it has good characteristics for accelerator control problems. It is highly sample-efficiency  and its representational power is sufficient for most common cases\cite{Kain2020,Hirlaender2020a}. Modifications to increase the stability are subject of \cref{ss:Normalized advantage function}.

@@

		V^\policy(\bs_t) := \mathbb E_{\traj \sim p_{\policy}(\traj|\bs_t)}\left[\sum_{t=t'}^{\horizon}\discount^{(t'-t)} r(\bs_t,\ba_t)\right].\label{eq:state-value-function}

@@

	The state-action-value function - short Q function - $Q^\policy(\bs,\ba)$, which tells us how good an action $\ba$ in a state $\bs$ following $\policy$ is, in terms of the expected return:

@@

		Q^\policy(\bs_t,\ba_t) := \mathbb E_{\traj \sim p_{\policy}(\traj|\bs_t,\ba_t)}\left[\sum_{t=t'}^{\horizon}\discount^{(t'-t)} r(\bs_t,\ba_t)\right],

@@

	is expressed as $Q^\policy_\theta(\bs,\ba)$, where $\theta$ denotes the parameters of the function approximator, such as the weights of a neural network. By satisfying the Bellmann-optimality equation $Q^\policy_\theta$ can be trained towards the optimal $Q^*(\bs,\ba)$ by minimizing the Bellman error\footnote{$\vec{Q}_\theta$ denotes the Q-function $Q_\theta$ represented as a vector of length $|\states|\times |\actions|$.}:

@@

		\policy_\theta(\bs_t)=\delta(\ba_t-\arg\!\max_\ba Q_\theta(\bs_t,\ba_t)).

@@

	The Bellman operator $\bellman^*$ has a unique fixed point but is non-linear, due to the involved $\max$ - operator \cite{Sutton2018}:

@@

	The nature of this equation can cause overestimation and other complications, when using a function approximator and several attempts exist to overcome its difficulties \cite{Hasselt2015,Mnih2013,Lillicrap2015,Gu2016,Wang2015}.
	\subsubsection{Design decisions for MFRL}\label{ss:Normalized advantage function}

@@

	If a specific quadratic form of the Q-function is assumed \cite{Gu2016}:

@@

		&\mathcal L(\theta)=\\\notag&\left( (\reward(\bs_t,\ba_t)+\gamma \min_{1,2} V_{\theta^i_\text{targ}}(\bs_{t+1}) - (1+\gamma) Q_\theta(\bs_t,\ba_t)\right)^2.

@@

	$\theta_\text{targ}$ are the weights of a target network, which is softly updated \cite{Lillicrap2015, Gu2016,Silver2014}. To further stabilize the network training a small clipped artificial noise is added to the actions called \emph{smoothing} \cite{fujimoto2018addressing}. The effect is demonstrated in \cref{appendix:naf2} on a classical example from control theory, the \emph{inverted pendulum}. In \cref{fig:comparsion_noise} its role to increase the noise robustness can be seen. Therefore \emph{smoothing} was applied in all MFRL experiments.\\
	 Our implementation has a high sample-efficiency and noise resistivity. We use it as the baseline for the FERMI-FEL control problem, as it yields good results. Additional changes to the original proposal \cite{Gu2016} and a previous implementation used for accelerator control using to a prioritized replay buffer \cite{Hirlaender2020a} are discussed in \cref{appendix:naf2}. We refer to our modified implementation as \emph{NAF2} and the implementation close to the original, as in \cite{Gu2016}, as \emph{NAF}.

@@

	A nice overview of state of the art MBRL methods is provided in \cite{Wang2019}. Suitable methods are \cite{Gal2016,6654139}, based on Gaussian processes and Bayesian Neural Networks and back-propagation in the dynamics model or \cite{Chua2018,Wang2019a}, using highly efficient sample based methods (e.g. the cross entropy method \cite{Boer2005}). We focus on model-based data generation.\\
		In MFRL the dynamics $\transitions$ of an environment is learned implicitly, while in model-based reinforcement learning MBRL $\transitions$ is learned explicitly and approximated as $\hat \transitions$. In addition to the next state, the reward is included in the model \footnote{In some applications $\reward$ can be deduced from the current state or the state change. In such a situation the modelling can be simplified to model directly $\hat\transitions$}:

@@

	A schematic overview of our used method is shown in \cref{fig:MBRL_overview}. Initially, data is gathered by interaction with the real environment or read from a previous collection. An uncertainty aware model of the dynamics is trained, using 'anchored ensembling' \cite{Pearce2018} on the data (details provided in \cref{appendix:The impact of noise,ss:The uncertainty aware dynamics model}), which allows taking the aleatoric (measurement errors) as well as the epistemic uncertainty (lack of data) into account. It was motivated by the fact that pure epistemic ensemble techniques as \cite{Kurutach2018} were reported to be very sensitive to aleatoric noise \cite{Wang2019}. Alternatives as \cite{Chua2018,Janner2019,Wang2019a} also take these uncertainties into consideration using probabilistic ensembles. Our implementation is simpler while showing excellent performance in considered experiments.\\
	 Subsequently, an MFRL algorithm is deployed to learn the controller on the dynamics model by only using the synthetic data and interoperating the model uncertainty, as explained in \cref{ss:The uncertainty aware dynamics model}.\\
	  After a defined number of training steps, called epoch, the controller is validated on each model of the ensemble, individually. If there is no improvement in a number smaller than the total number of the models, the controller might be tested on the real environment for several episodes. In this way 'over-fitting' on a wrong model is avoided, as discussed in \cite{Kurutach2018}. The training is stopped if the controller produces satisfying results on the real environment. Otherwise, a batch of new data is collected and the model is improved and consecutively a new epoch starts.

@@

	\subsection{Critical design decisions in MBRL}\label{ss:critical_design}
In the following, the most important aspects of a successful application of MBRL are discussed. To our knowledge uncertainty aware MBRL is applied for the first time with success on an accelerator problem and might be, despite its incredible potential, demanding. We made some algorithmic design decisions, which we think are beneficial for problems typically found in accelerator operation.\\

@@

\subsubsection{The uncertainty aware dynamics model}\label{ss:The uncertainty aware dynamics model}
The 'anchored ensembling' method usually yields good results already with a small number of models. Empirical results showed that already three models were satisfactory to see definite improvements over a single network approach (see \cref{fig:Compare_models_sizes}). The main goal is not to determine the exact posterior probability of the model, but, as already mentioned, providing the uncertainty to not over-train the controller at areas without sufficient data.\\

@@

The prior is controlled by the initialisation of the weights of the network. It defines how large the variation of the approximated function is expected and were randomized uniformly in a small interval $[-\Delta, \Delta]$. The weights of the last layer were normalised by the number of its nodes. For the moment, only homoskedastic Gaussian errors are considered with a standard deviation is $\sigma_\epsilon$ is also respected in a regularisation term \cref{appendix:The impact of noise} and significantly improves the training in the presence of noise, as shown in \cref{fig:comparsion_noise_ae_dyna}. \\
Several methods of network training were implemented. \emph{Early stopping} \cite{Goodfellow2016}, with a learning rate of $10^{-3}$ to $10^{-4}$  was employed  with a waiting time of about 20 steps and a validation ratio of 20\%. The number of training steps is increasing with more data in close-by areas. It is followed by a shrinking of the uncertainty at regions where no data is available, leading to better training performance. A fixed maximal loss threshold was also beneficially tested. In the experiments, a combination of both was taken.
\subsubsection{The controller algorithm}
To decide which MFRL algorithm to use, two already mentioned main algorithm classes are considered  (\cref{s:Model free reinforcement learning}): \emph{on-} and  \emph{off-policy} algorithms \cite{Sutton2018}. \emph{On-policy} algorithm show a stabler convergence in general, while \emph{off-policy} algorithms have the advantage that the data buffer can be filled with real data.\\
The \emph{on-policy} algorithm trust region policy optimization (TRPO) \cite{Schulman2015} provides the theoretical guarantee of a monotonic improvement of the policy. We use this algorithm in experiments labelled as \emph{ME-TRPO}, which stands for model-ensemble TRPO as proposed in \cite{Kurutach2018}. As reported in \cite{Kurutach2018}, using proximal policy optimisation \cite{Schulman2017}, another prominent \emph{on-policy} algorithm, reveals minor performance, which was confirmed in our tests. \\
An attractive off-policy algorithm is the soft-actor-critic (\emph{SAC}) algorithm. \cite{fujimoto2018addressing,Hill2018}. \emph{SAC} not only tries to maximise \cref{eq:cumulative_reward} but also simultaneously is regularised to maximise the entropy in the action space \cite{Haarnoja2018a}. In this way, exploration is encouraged to avoid getting stuck in a local optimum and a good trade-off between exploration and exploitation is achieved. We refer to experiments using the \emph{SAC} as \emph{AE-DYNA-SAC}. 
For \emph{AE-DYNA-SAC} the controller was reset after each retraining of the dynamics model, to profit from its exploration features, while for \emph{ME-TRPO} the policy was constantly improved to exploit its monotony.
\subsubsection{Handling of the uncertainty}
One of the most crucial points is how model uncertainties are taken into account by the MFRL controller. Several different approaches were tried. One way is to take the mean of the models $\mu_m$ and the standard deviation $\sigma_m$ and sample from $\gauss(\mu_m,\sigma_m)$. Another strategy is to randomly select a model to provide at each training step, which demonstrated the best performance the original implementation of the model ensemble \emph{TRPO} (\emph{ME-TRPO}) algorithm and was used in the experiments labelled \emph{ME-TRPO} \cite{Kurutach2018}. A pessimistic selection would only select the model resulting in the lowest predicted reward (as used e.g. in \cite{kidambi2020morel}) and showed constant but too slow improvement for our online-training.
Good results were also obtained by following a randomly selected single model each full episode in combination, which was used in the \emph{AE-DYNA-SAC}. 
\subsubsection{The data acquisition}
The number of data-points in a new batch to improve the model has to be chosen carefully. It is influenced by the number of randomly collected initial data-points (and then by the used policy $\policy$). The initialisation phase has to be selected not to small to minimise risks getting trapped in a local minimum for too long. Afterwards, from our experience, at least one full episode per epoch should be taken. We decided to use a short horizon to diminish the impact of the compound error \cite{Janner2019}, the accumulation error following a wrong model, as well known in DYNA-style approaches. The maximal number of steps in our experimental runs per episode was ten. During the data acquisition on the real system, the latest learned policy $\policy$ was taken with some small additional Gaussian noise to improve exploration.
%\tocheck{Niky discuss the averaging of the data}
	\section{Experimental results from FERMI RL online tests}\label{sec:Experimental results from FERMI RL online tests}
	As discussed in the previous sections, several tests were performed on the FERMI FEL.
	The main purpose was to try the newly implemented algorithms on a real system to evaluate their operational feasibility. 
	Because of the tight schedule at FERMI, only a few shifts of several hours could be reserved to carry out these experiments.\\ 
%	Since there was a positive outcome, we deduce that the utilised methods are at an operational level.\\
